---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

We will import this as a CSV file and leave it as is for this analysis. 

```{r}
activitydata<-read.csv("activity.csv")
head(activitydata)
summary(activitydata)
```

## What is mean total number of steps taken per day?

Here is a histogram for the total number of steps taken each day:

```{r}
totaldailysteps<-tapply(activitydata$steps,activitydata$date,sum,na.rm=TRUE)
hist(totaldailysteps, xlab = "Total Daily Steps", main = "Histogram of Total Daily Steps")
```

Here is the median of the total number of steps taken each day:

```{r}
allmedians<-tapply(activitydata$steps,activitydata$date,median,na.rm=TRUE)
summary(allmedians)
```

Here is the mean of the total number of steps take each day:
```{r}
allmeans<-tapply(activitydata$steps,activitydata$date,mean,na.rm=TRUE)
summary(allmeans)
```
## What is the average daily activity pattern?

Fist, we will extract the average number of steps per interval. Then we will plot the daily activity pattern.

```{r}
intervalmean<-tapply(activitydata$steps,activitydata$interval,mean,na.rm=TRUE)
plot(x=dimnames(intervalmean)[[1]],y=intervalmean,type="l", main = "Daily Activity Pattern", xlab = "Intervals - 5 Minutes", ylab = "Interval Means")
```

## Imputing missing values

There seems to be a notable number of missing values. Let's take a look:
```{r}
sum(is.na(activitydata$steps))
```


## Are there differences in activity patterns between weekdays and weekends?
